Reachability analysis for continuous one-counter automata

### 2021-08-21 Documentation/paper.tex
updated documentation

\documentclass[12pt]{article}

@@

\cleardoublepage <*-+/

@@

\phantomsection
\addcontentsline{toc}{section}{Nederlandstalige samenvatting}
\section*{Nederlandstalige samenvatting}
Een gekend probleem bij het schrijven van code is het vinden van lijnen code die niet bereikbaar zijn \cite{713323, 10.1145/1292316.1292319}. Onbereikbare code kan veel oorzaken hebben maar één van de moeilijker te analyseren situaties zijn condities die nooit voldaan worden. In deze gevallen is de code omringd door de conditie niet bereikbaar en zou deze dus eigenlijk weg moeten.

@@

Niet alleen de functie inhoud maar ook de definitie zal een beperking krijgen, zo accepteren we enkel parameters van het type integer.
\newpage
\section{Introduction}
A well known problem in writing code is determining whether or not every line of code is reachable. It is often impossible to properly determine whether or not a line of code guarded by a conditional statement is going to be reachable in any possible execution of the code \cite{713323, 10.1145/1292316.1292319}. Unreachable code is never desired and is an ever more pressing issue for energy restrained environments such as embedded computer systems \cite{embeddedefficiency}. In these systems, we want our execution to be as efficient as possible, and therefore avoid conditions that can never be satisfied or lines that can never be reached. To remedy this, a proper unreachable code detection approach is needed.

@@

\subsection{Research questions}
The objective of this research can be summarised in two core research questions:

@@

\section{Motivation}
Code efficiency is one of the core aspects on which the quality of code is measured. Inefficient code can incrementally lead to a slower product, even though it only represents a small part of a much larger whole. This is because code reuse is a critical aspect in Object Oriented Programming (OOP) \cite{10.1145/71605.71612}. In OOP we tend to write code by encapsulating it within classes (objects) based on the functionality they provide. This allows us to simply reuse that object whenever we need that specific functionality. As a direct result of this, every time we introduce code inefficiency, this will get replicated whenever we need use of the specific functionality in which the inefficiency was introduced.

@@

As can be concluded from the examples given above, there has been research on dead code analysis. This research is however dated and it did not specifically cover the environment which we will cover within this research. It is furthermore shown that within the majorly used IDE's such as Eclipse and Clion the cases which we will attempt to detect with this new approach are not yet covered. We can conclude that this research brings additional value to what is known about dead code analysis and can further enhance the optimization of software.
\section{Preliminaries}
This paper will make use of counter automata and more specifically continuous one counter automata. These automata are derived from the general definition of automata and signify an extension by adding the notion of counters.

\subsection{General automata}
The automata used in this research are a specific subset of the generally defined automata. The general definition for an automaton A is given by the tuple (Q, $\Sigma$, $\delta$, $q_0$, F) where
\begin{itemize}
	\item Q is the set of states
	\item $\Sigma$ is a finite set of symbols, called the alphabet of the automaton
	\item $\delta: Q \times \Sigma \rightarrow Q$ is the transition function
	\item $q_0 \in Q$ is the initial state
	\item $F \subseteq Q$ is a set of states of Q defining the final states, 
\end{itemize}

\subsection{Continuous one-counter automata}
A continuous one-counter automaton (COCA) is defined as a tuple $\mathcal{A}$ = $(Q, q_0, T, \tau)$ where
\begin{itemize}
	\item $Q$ is a finite set of states
	\item $q_0 \in Q$ is the initial state
	\item $T \subseteq Q \times op \times Q$ is the transition relation with $op$ the set of operations
	\item $I$ is a representation of all intervals over $\mathbb{Q}$
	\item $\tau: Q \rightarrow I$ is the mapping of the reachability intervals
\end{itemize}

@@

\subsection{COCA reachability}
In this section we give an overview of how we approach the concept of reachability. The provided concept was proven by Michael Blondin et al. \cite{blondin2021continuous} and will be the basis for the remainder of the approach.

@@

\section{Analysis of dead code in C}
The approach given within this paper consists out of two major phases. The first phase will consist of the conversion from C code to one counter automata. The second phase will use the result from the first phase, and perform a reachability analysis on these generated automata. The first phase was already covered in preceding research and the majority of the notes on the approach in the first phase were taken from that research. They were copied here as this research will conclude the unfinished work that was done there, as well as make continuous references to the flaws which arose from it.

@@

\subsection{Preprocessing}
\label{sec:conversion}
The program will start by resolving all the compiler macros that are part of C. These are intended to be resolved by the compiler at compile time, and are often dependent on which compiler was chosen/the environment on which the code is executed. For our use case I decided to select the global variables in such a way that these represent a windows machine that is using 64 bits. This is however not at all restricted to that and could further be expanded to support any environment and any combination of global variables possible. This selection is in essence not relevant as it should not affect the presence of dead code.

@@

\subsection{Initialization}
Using the flattened result of the preprocessor, the next step will start by running the generator ANTLR compilation unit on the given C code. In case there are any issues regarding simple compilation rules, the program will exit with an error, indicating the issues in the code. For the remainder of the execution, the code is assumed to be correct, and there will be no regard for potential issues within the code itself.

\subsection{Context-free abstract syntax tree generation}
Using the parse tree resulting from the ANTLR compilation, we will try to generate an abstract syntax tree that is as simple as possible without regarding any context.

@@

However, the creation of nodes is not the same in all situations, in some cases, a specific kind of node will be created and in other cases it will not, depending on the rule used for the current node (which we can determine from the context of the current node). In order to know whether or not we need to pop a node from the node stack, we will allow the code to look at the top node in the stack, and evaluate the type of set node. No other context related evaluations will be allowed.\\
\\
\noindent
\textbf{Context-free analysis assignmentExpression example}\\
To look at a concrete example, consider the following rule in the grammar:
\begin{align*}
	&assignmentExpression\\
	&\text{:}\indent \text{conditionalExpression}\\
	&|\indent \text{unaryExpression assignmentOperator assignmentExpression}\\
	&|\indent \text{Digitsequence}; \\
\end{align*}
The rule for conditionalExpression needs to be there, as there is a chain of expressions, and some expressions need to be higher in the order than others, therefore all expressions with a higher priority need to be evaluated before the lower priority expressions. This is done by adding a rule to a lower priority expression in each expression class. It is however undesirable for these nodes to be added. Without regard for context in other nodes, we can simply look at the current node to see whether or not the conditionalExpression line is used (which ANTLR generated context allows us to do).

@@

\subsection{Context-sensitive abstract syntax tree reduction}
There is still the need for reductions. We will attempt to implement constant propagation and substitution, in which we compute constant expressions, and replace variables with known value, by their actual value. Furthermore we will try to rephrase expressions by expressions with the same meaning, so that all expressions, if possible, are in an acceptable format for the counter automaton.

@@

\subsection{Abstract syntax tree validation}
The fourth loop will iterate over the abstract syntax tree resulting from the previous step, and while doing so, it will evaluate the nodes that occur and the context in which they occur.

@@

Important to note is that there is no regard for global statements. These are considered to be variable and undetermined. It could be that they can perfectly be used as counters, and that they will never be altered by other functions, but it is impossible to determine this efficiently. Therefore, a counter must be declared within the function scope.\\
\\
\noindent
\textbf{Validation example}\\
When we go back to the earlier mentioned segment of code, given in Listing \ref{lst:ready to eval}, we obtain the automaton displayed in Figure \ref{fig:divide_cleaned}.

@@

The selection statement will be evaluated in the exact same manner as the Iteration Statement was evaluated, and as the jump statement holds no special operations, we can conclude that this function satisfies the requirements for continuous one counter automaton generation.

\subsection{Counter automaton generation}
\label{sec:generation}
The final step of the conversion will be the counter generation itself. In this loop, there will be no more regards to possible invalid statements. At this point we assume everything to be known, and we can therefore simplify our analysis. Recall that the following list is the collection of allowed labels for the counter automaton.
\[
op = \{+c, -c, +p, -p, \leq c, =c, \geq c, \leq p, =p, \geq p: c \in \mathbb{N}, p \in P\} \cup \{\wedge\}
\]
This allows us to just check the head nodes related to such expressions.

\subsubsection{Functions}
First of all, we need the notion of the \textit{Function Definition} statements. The generator will generate a counter automaton for each function, and needs to know when these start.

\subsubsection{Assignments}
Other than functions, we will also need to know when assignments to counters occur, so that we can add the proper transition labels. We only need to check whether or not we are assigning to a counter, if so, we read the operation used, and the variable that gets assigned. Valid counter assignments need to be of the form \textit{c op x} where $c$ is the counter, $op \in $ \textit{\{=, +=, -=\}} and $x \in P \lor x \in \mathbb{N}$.

@@

\subsubsection{Inequality conditions}
\label{sec:inequality}
Another statement we will need to support, are inequality conditions. If we want our conditional statements to be properly functioning, we will need to be able to generate opposing conditions, but the opposing condition of the equality condition, is the inequality condition, which we can not directly model using the given operations.

@@

\subsubsection{Less than and greater than}
\label{sec:strict}
The final addition to the set of supported statements, are the \textit{less than} and \textit{greater than} expressions. These are needed to model the opposing conditions to greater than or equal and less than or equal respectively.

@@

\subsubsection{Iteration statements}
We will also need the notion of iteration statements, which needs 5 additional special nodes to fully support it's functionality. The first node will be the \textit{start of the loop}. The second node is the \textit{pre node} which symbolizes the state after the precondition of the for loop has been evaluated. Next we need \textit{start of the inner segment} to signify the start of the inner segment. This node will be reached by a transition with the condition as a label.

@@

Consider the segment of code in Listing \ref{lst:itteration} for which the code conversion will result in the automaton in Figure \ref{fig:test_counter_automaton}.
\begin{lstlisting}[style=CStyle, caption={Example of an itteration statement}, label={lst:itteration}]
bool test(int c){
	for(c = 0; c < 5; c ++){
		continue;
	}
	return true;
}
\end{lstlisting}

@@

\subsubsection{If statements}
\label{sec:if}
Another important construct are the if statements. These statements will, just as the iteration statements did, require additional nodes to be able to be modelled. If statements require a total of 4 nodes to be modelled.

First of all, we need a $start\ if$ node, from which the conditions will start. The $if$ and $else$ nodes will be used to signify the start of the if and else segment (if present). Independent of the presence of an $else$ node the if configuration will always end in an $end\ if$ node from which point the automaton will carry on to the next line of code.

@@

@@

\subsubsection{Switch statements}
A second conditional statement that exists in C, is a \textit{switch statement}. This statement allows a variable to be compared against multiple values. What is special about the switch statements, is that if matched with one case, the code of all consequential cases will also be executed, until the end of the switch, or until a jump occurs.

@@

\subsubsection{Counter initialization}
The counters do not get initialized at the point that they got declared within the code. This is a simple improvement that helps to reduce the number of counters. Consider the code segment in Listing \ref{lst:two counters}, in which two different counters are used.

@@

\subsubsection{Final overview}
To give an in depth overview of how the conversion works, we will transform the following code snippet given in Listing \ref{lst:final overview} into a counter automaton. This code consists of multiple if statements, as well as a loop, it makes use of a single counter value, being \textit{counter}, and two parameters. The loop and the final \textit{if} both use inequality expressions, which are representable in a counter automaton, as was described earlier. The first two \textit{ifs} make use of relational expressions with the less than and greater than operators, which are also representable within a counter automaton. The counter is only  using additions or subtractions with parameters and none of these parameters will ever be modified, which also conforms to the requirements. Finally, this function has only integer parameters which makes us conclude that this entire function is representable in a counter automaton.
\newpage

@@

\subsection{Automaton specification format}
\label{sec:automaton input}
For the approach described within this paper, the dot language part of Graphviz\cite{10.1007/3-540-45848-4_57} was assumed to be used to represent all automata textually. This language is a very versatile and easily understandable format in which automata can be defined. It was also used as the output language in the preceding research which automatically made it the most interesting format to support. No other formats were considered as the conversion from a textual representation to an instance of the \textit{Automaton} class is not considered to be the core of the research. The logic used for the dot language is furthermore identical to the logic that would need to be applied to read any other automaton specification, with the only difference lying in the language specific aspects of the specification.

\begin{landscape}
	\begin{figure}
		\centering
		\includegraphics[width=\linewidth]{final_overview_11}
		\caption{Overview of the final resulting counter automaton}
		\label{fig:final_overview_11}
	\end{figure}
\end{landscape}

\subsubsection{Edge transformation}
Edges are represented within the dot language by a single line per edge using the format specified below. The \textit{origin} and \textit{end} tokens specify the start and end node of the transition. The square brackets encapsulate optional arguments, of which the label and xlabel arguments matter most, as either can be used to specify the label of the transition, and thereby the operation that we want to have attached to the edge.

@@

To closely match the specification of the theoretical proof on which this approach is based \cite{blondin2021continuous}, we require edges to only contain operations and not conditions. In case an edge is encountered with a condition as a label, this will be resolved by introducing a new node, \textit{inbetween}, to which this condition will be added. This node will be added as an intermediary step and the condition label will be attached to this node. By ensuring that this node will only have one incoming edge and one outgoing edge we preserve the fact that the condition must hold in case one desires to from from the \textit{origin} node to the \textit{end} node.

@@

\subsubsection{Node transformation}
Nodes do not require to be specified using a separate node specification rule, and will simply be generated in case they are used within an edge. Node specifications function specifically for the cases in which additional attributes need to be added to nodes. To support operations as well as general labels, the only two attributes which will be tracked from node specification are the \textit{label} and \textit{xlabel} arguments. These labels will be matched with regexes specifying the supported formats for operations and conditions. In case the label does not match either, it is still tracked as regular node labels are still relevant for reachability reporting. For the approach discussed within this paper the node labels represent the code line from which they were generated and are therefore needed to analyse code line reachability. All node specifications will follow the format given below where the square bracket can contain a number of optional arguments.

@@

\subsubsection{Initial node}
For reachability analysis to be relevant there needs to be an initial node. In automata, the initial node is often represented by a transition going from nowhere towards a single node. This is however not a built in feature within the dot language. One of the solutions to overcome this is to create a node and make it invisible, an edge can then be drawn from this invisible node to the actual initial node. Other approaches are potentially possible but were not explored. Any alternative implementations could easily be introduced instead of the considered format without interchanging any of the other components.

@@

\subsection{Reachability of non-parametric COCA}
\label{sec:reachability non parametric}
In this section we will give a description of the solution that was used to determine the reachability for non-parametric continuous one counter automata. This will be done by showing the approach that was taken, as well as an application to show its usage. This approach is a practical implementation of the theoretical approach described by Michael Blondin et al. \cite{blondin2021continuous}.

\subsubsection{Initialization}
During the initialization phase, we will instantiate all reachability intervals. The initial node will get an initial reachability interval $[i, i]$ where $i$ represents the initial counter value. This is not a fixed value, but will for the sake of this research always default to 0, simply because all integer values in most programming languages default to 0 at initialization.

@@

\subsubsection{Interval updates}
\label{interval updates}
From the initial state, we will start doing cycles in which we update all intervals once, wherever possible. This will be done using the successor function \textit{succ} below where $R_i: Q \times Q \rightarrow 2^\mathbb{Q}$ represent the mapping function for the i-th iteration. This function is going from each transition to its corresponding counter reachability interval. The \textit{succ} function will update the interval of one transition, \textit{(p, z, q)}, to its next iteration, based upon the current configuration.

@@

\subsubsection{Loop acceleration}
\label{sec:accel}
Unfortunately, the \textit{succ} function is not guaranteed to find the bounds in a finite number of evaluations. For example, the scenario given in Figure 23 would never converge at all. The interval of the loop between $q_1$ and $q_2$ will keep expanding indefinitely and the evaluation will therefore never stop.

@@

\subsubsection{Proof of termination}
The mathematical proof of termination has already been given by Michal Blondin et al. \cite{blondin2021continuous} but an intuitive proof for the algorithm will also be given here. Given that there are no loops it will trivially terminate within n evaluations, given that n is equal to the total number of nodes within the automaton. If there are loops they must either be positively expanding, negatively expanding, expanding in both directions or not expanding at all.

@@

\subsection{Reachability of parametric COCA}
\label{sec:reachability parametric}
Throughout this section we will give an overview of how the reachability for parametric continuous one counter automata can be analysed. This will be done using the approach that was theoretically described by Michael Blondin et al. \cite{blondin2021continuous}. This paper will both give a practical approach as well as an application, intended as prove that it works in practise.

@@

\subsubsection{Transition initialization}

@@

\subsubsection{Node condition initialization}

@@

\subsubsection{Interval encoding}

@@

\subsubsection{Define the end goal}
One by one we will verify for each node whether or not it is reachable, this can be done using the condition below, where e represents the edge taken to generate any of the intervals part of the goal node. In case e is equal to -2 it signifies that no transition was taken and that the interval is empty. If e is not equal to -2 it must imply that the according interval is not empty as will be explained later.

@@

\subsubsection{msum}
The msum function is used to determine the inclusiveness of the bounds in the interval addition. The msum function is defined as:

@@

\subsubsection{Emptiness of intervals}
We need to be able to verify whether or not an interval is empty in order to create base cases for future operations. An interval is considered to be empty if both bounds are not inclusive and the boundary values are equal to each other as can be seen in the formulae below.

@@

\subsubsection{Interval addition}
The addition of intervals will be split in three different cases as can be seen in the graph below. In case either operand is an empty interval the resulting interval will trivially equal the other interval. In case they are both not empty we compute the sum by taking the addition of the lower and upper bound of the X and Y intervals and the msum to compute whether or not the resulting bounds are inclusive. An overview of this can be found in Figure \ref{fig:addition}.

@@

		\draw [arrow] (q1) |- (q4);
		\draw [arrow] (q2) |- (q5);
		\draw [arrow] (q3) |- (q6);
	\end{tikzpicture}
	\caption{Overview of the interval addition operation}
	\label{fig:addition}
\end{figure}

\subsubsection{Interval overlap}
\begin{figure}[b!]
	\centering
	\begin{tikzpicture}[auto, >=latex, font=\footnotesize\scshape]
		\tikzstyle{small} = [draw=black, rectangle]
		\tikzstyle{center} = [draw=black, rectangle, text width=8em, align=center]
		\tikzstyle{arrow} = [thick,->,>=stealth]

@@

\subsubsection{Interval union}
The union operation is summarised in Figure \ref{fig:union} below. We can immediately see that this is very similar to the intersection in the sense that most of the conditional statements are the exact same.

>>>>

\documentclass[12pt]{thesis}

@@

@@

\phantomsection
\addcontentsline{toc}{section}{Nederlandstalige samenvatting}
\chapter*{Nederlandstalige samenvatting}
Een gekend probleem bij het schrijven van code is het vinden van lijnen code die niet bereikbaar zijn \cite{713323, 10.1145/1292316.1292319}. Onbereikbare code kan veel oorzaken hebben maar één van de moeilijker te analyseren situaties zijn condities die nooit voldaan worden. In deze gevallen is de code omringd door de conditie niet bereikbaar en zou deze dus eigenlijk weg moeten.

@@

Niet alleen de functie inhoud maar ook de definitie zal een beperking krijgen, zo accepteren we enkel parameters van het type integer.
\newpage
\chapter{Introduction}
A well known problem in writing code is determining whether or not every line of code is reachable. It is often impossible to properly determine whether or not a line of code guarded by a conditional statement is going to be reachable in any possible execution of the code \cite{713323, 10.1145/1292316.1292319}. Unreachable code is never desired and is an ever more pressing issue for energy restrained environments such as embedded computer systems \cite{embeddedefficiency}. In these systems, we want our execution to be as efficient as possible, and therefore avoid conditions that can never be satisfied or lines that can never be reached. To remedy this, a proper unreachable code detection approach is needed.

@@

\section{Research questions}
The objective of this research can be summarised in two core research questions:

@@

\chapter{Motivation}
Code efficiency is one of the core aspects on which the quality of code is measured. Inefficient code can incrementally lead to a slower product, even though it only represents a small part of a much larger whole. This is because code reuse is a critical aspect in Object Oriented Programming (OOP) \cite{10.1145/71605.71612}. In OOP we tend to write code by encapsulating it within classes (objects) based on the functionality they provide. This allows us to simply reuse that object whenever we need that specific functionality. As a direct result of this, every time we introduce code inefficiency, this will get replicated whenever we need use of the specific functionality in which the inefficiency was introduced.

@@

As can be concluded from the examples given above, there has been research on dead code analysis. This research is however dated and it did not specifically cover the environment which we will cover within this research. It is furthermore shown that within the majorly used IDE's such as Eclipse and Clion the cases which we will attempt to detect with this new approach are not yet covered. We can conclude that this research brings additional value to what is known about dead code analysis and can further enhance the optimization of software.

\chapter{Preliminaries}
This paper will make use of counter automata and more specifically continuous one counter automata. These automata are derived from the general definition of automata and signify an extension by adding the notion of counters.

\section{General automata}
The automata used in this research are a specific subset of the generally defined automata. The general definition for an automaton A is given by the tuple (Q, $\Sigma$, $\delta$, $q_0$, F) where
\begin{itemize}
	\item Q is the set of states
	\item $\Sigma$ is a finite set of symbols, called the alphabet of the automaton
	\item $\delta: Q \times \Sigma \rightarrow Q$ is the transition function
	\item $q_0 \in Q$ is the initial state
	\item $F \subseteq Q$ is a set of states of Q defining the final states, 
\end{itemize}

\section{Continuous one-counter automata}
A continuous one-counter automaton (COCA) is defined as a tuple $\mathcal{A}$ = $(Q, q_0, T, \tau)$ where
\begin{itemize}
	\item $Q$ is a finite set of states
	\item $q_0 \in Q$ is the initial state
	\item $T \subseteq Q \times op \times Q$ is the transition relation with $op$ the set of operations
	\item $I$ is a representation of all intervals over $\mathbb{Q}$
	\item $\tau: Q \rightarrow I$ is the mapping of the reachability intervals
\end{itemize}

@@

\section{COCA reachability}
In this section we give an overview of how we approach the concept of reachability. The provided concept was proven by Michael Blondin et al. \cite{blondin2021continuous} and will be the basis for the remainder of the approach.

@@

\chapter{Analysis of dead code in C}
The approach given within this paper consists out of two major phases. The first phase will consist of the conversion from C code to one counter automata. The second phase will use the result from the first phase, and perform a reachability analysis on these generated automata. The first phase was already covered in preceding research and the majority of the notes on the approach in the first phase were taken from that research. They were copied here as this research will conclude the unfinished work that was done there, as well as make continuous references to the flaws which arose from it.

@@

\section{Preprocessing}
\label{sec:conversion}
The program will start by resolving all the compiler macros that are part of C. These are intended to be resolved by the compiler at compile time, and are often dependent on which compiler was chosen/the environment on which the code is executed. For our use case I decided to select the global variables in such a way that these represent a windows machine that is using 64 bits. This is however not at all restricted to that and could further be expanded to support any environment and any combination of global variables possible. This selection is in essence not relevant as it should not affect the presence of dead code.

@@

\section{Initialization}
Using the flattened result of the preprocessor, the next step will start by running the generator ANTLR compilation unit on the given C code. In case there are any issues regarding simple compilation rules, the program will exit with an error, indicating the issues in the code. For the remainder of the execution, the code is assumed to be correct, and there will be no regard for potential issues within the code itself.

\section{Context-free abstract syntax tree generation}
Using the parse tree resulting from the ANTLR compilation, we will try to generate an abstract syntax tree that is as simple as possible without regarding any context.

@@

However, the creation of nodes is not the same in all situations, in some cases, a specific kind of node will be created and in other cases it will not, depending on the rule used for the current node (which we can determine from the context of the current node). In order to know whether or not we need to pop a node from the node stack, we will allow the code to look at the top node in the stack, and evaluate the type of set node. No other context related evaluations will be allowed.\\
\newpage
\noindent
\textbf{Context-free analysis assignmentExpression example}\\
To look at a concrete example, consider the following rule in the grammar:
\begin{align*}
	&assignmentExpression\\
	&\text{:}\indent \text{conditionalExpression}\\
	&|\indent \text{unaryExpression assignmentOperator assignmentExpression}\\
	&|\indent \text{Digitsequence}; \\
\end{align*}
The rule for conditionalExpression needs to be there, as there is a chain of expressions, and some expressions need to be higher in the order than others, therefore all expressions with a higher priority need to be evaluated before the lower priority expressions. This is done by adding a rule to a lower priority expression in each expression class. It is however undesirable for these nodes to be added. Without regard for context in other nodes, we can simply look at the current node to see whether or not the conditionalExpression line is used (which ANTLR generated context allows us to do).

@@

\section{Context-sensitive abstract syntax tree reduction}
There is still the need for reductions. We will attempt to implement constant propagation and substitution, in which we compute constant expressions, and replace variables with known value, by their actual value. Furthermore we will try to rephrase expressions by expressions with the same meaning, so that all expressions, if possible, are in an acceptable format for the counter automaton.

@@

\section{Abstract syntax tree validation}
The fourth loop will iterate over the abstract syntax tree resulting from the previous step, and while doing so, it will evaluate the nodes that occur and the context in which they occur.

@@

Important to note is that there is no regard for global statements. These are considered to be variable and undetermined. It could be that they can perfectly be used as counters, and that they will never be altered by other functions, but it is impossible to determine this efficiently. Therefore, a counter must be declared within the function scope.\\
\newpage
\noindent
\textbf{Validation example}\\
When we go back to the earlier mentioned segment of code, given in Listing \ref{lst:ready to eval}, we obtain the automaton displayed in Figure \ref{fig:divide_cleaned}.

@@

The selection statement will be evaluated in the exact same manner as the Iteration Statement was evaluated, and as the jump statement holds no

\noindent
special operations, we can conclude that this function satisfies the requirements for continuous one counter automaton generation.

\section{Counter automaton generation}
\label{sec:generation}
The final step of the conversion will be the counter generation itself. In this loop, there will be no more regards to possible invalid statements. At this point we assume everything to be known, and we can therefore simplify our analysis. Recall that the following list is the collection of allowed labels for the counter automaton.
\[
op = \{+c, -c, +p, -p, \leq c, =c, \geq c, \leq p, =p, \geq p: c \in \mathbb{N}, p \in P\} \cup \{\wedge\}
\]
This allows us to just check the head nodes related to such expressions.

\subsection{Functions}
First of all, we need the notion of the \textit{Function Definition} statements. The generator will generate a counter automaton for each function, and needs to know when these start.

\subsection{Assignments}
Other than functions, we will also need to know when assignments to counters occur, so that we can add the proper transition labels. We only need to check whether or not we are assigning to a counter, if so, we read the operation used, and the variable that gets assigned. Valid counter assignments need to be of the form \textit{c op x} where $c$ is the counter, $op \in $ \textit{\{=, +=, -=\}} and $x \in P \lor x \in \mathbb{N}$.

@@

\subsection{Inequality conditions}
\label{sec:inequality}
Another statement we will need to support, are inequality conditions. If we want our conditional statements to be properly functioning, we will need to be able to generate opposing conditions, but the opposing condition of the equality condition, is the inequality condition, which we can not directly model using the given operations.

@@

\subsection{Less than and greater than}
\label{sec:strict}
The final addition to the set of supported statements, are the \textit{less than} and \textit{greater than} expressions. These are needed to model the opposing conditions to greater than or equal and less than or equal respectively.

@@

\subsection{Iteration statements}
We will also need the notion of iteration statements, which needs 5 additional special nodes to fully support it's functionality. The first node will be the \textit{start of the loop}. The second node is the \textit{pre node} which symbolizes the state after the precondition of the for loop has been evaluated. Next we need \textit{start of the inner segment} to signify the start of the inner segment. This node will be reached by a transition with the condition as a label.

@@

Consider the segment of code in Listing \ref{lst:itteration} for which the code conversion will result in the automaton in Figure \ref{fig:test_counter_automaton}.

\begin{lstlisting}[style=CStyle, caption={Example of an itteration statement}, label={lst:itteration}]
bool test(int c){
	for(c = 0; c < 5; c ++){
		continue;
	}
	return true;
}
\end{lstlisting}

@@

\subsection{If statements}
\label{sec:if}
Another important construct are the if statements. These statements will, just as the iteration statements did, require additional nodes to be able to be modelled. If statements require a total of 4 nodes to be modelled.

First of all, we need a $start\ if$ node, from which the conditions will start.

@@

\noindent
The $if$ and $else$ nodes will be used to signify the start of the if and else segment (if present). Independent of the presence of an $else$ node the if configuration will always end in an $end\ if$ node from which point the automaton will carry on to the next line of code.

@@

\subsection{Switch statements}
A second conditional statement that exists in C, is a \textit{switch statement}. This statement allows a variable to be compared against multiple values. What is special about the switch statements, is that if matched with one case, the code of all consequential cases will also be executed, until the end of the switch, or until a jump occurs.

@@

\subsection{Counter initialization}
The counters do not get initialized at the point that they got declared within the code. This is a simple improvement that helps to reduce the number of counters. Consider the code segment in Listing \ref{lst:two counters}, in which two different counters are used.

@@

\subsection{Final overview}
To give an in depth overview of how the conversion works, we will transform the following code snippet given in Listing \ref{lst:final overview} into a counter automaton. This code consists of multiple if statements, as well as a loop, it makes use of a single counter value, being \textit{counter}, and two parameters. The loop and the final \textit{if} both use inequality expressions, which are representable in a counter automaton, as was described earlier. The first two \textit{ifs} make use of relational expressions with the less than and greater than operators, which are also representable within a counter automaton. The counter is only  using additions or subtractions with parameters and none of these parameters will ever be modified, which also conforms to the requirements. Finally, this function has only integer parameters which makes us conclude that this entire function is representable in a counter automaton.

@@

\section{Automaton specification format}
\label{sec:automaton input}
For the approach described within this paper, the dot language part of Graphviz\cite{10.1007/3-540-45848-4_57} was assumed to be used to represent all automata textually. This language is a very versatile and easily understandable format in which automata can be defined. It was also used as the output language in the preceding research which automatically made it the most interesting format to support. No other formats were considered as the conversion from a textual representation to an instance of the \textit{Automaton} class is not considered to be the core of the research. The logic used for the dot language is furthermore identical to the logic that would need to be applied to read any other automaton specification, with the only difference lying in the language specific aspects of the specification.

\subsection{Edge transformation}
Edges are represented within the dot language by a single line per edge using the format specified below. The \textit{origin} and \textit{end} tokens specify the start and end node of the transition. The square brackets encapsulate optional arguments, of which the label and xlabel arguments matter most, as either can be used to specify the label of the transition, and thereby the operation that we want to have attached to the edge.

@@

To closely match the specification of the theoretical proof on which this approach is based \cite{blondin2021continuous}, we require edges to only contain operations and not conditions. In case an edge is encountered with a condition as a label, this will be resolved by introducing a new node, \textit{inbetween}, to which this condition will be added. This node will be added as an intermediary step and the condition label will be attached to this node. By ensuring that this node will only have one incoming edge and one outgoing edge we preserve the fact that
\begin{landscape}
	\begin{figure}
		\centering
		\includegraphics[width=\linewidth]{final_overview_11}
		\caption{Overview of the final resulting counter automaton}
		\label{fig:final_overview_11}
	\end{figure}
\end{landscape}

\noindent
the condition must hold in case one desires to from from the \textit{origin} node to the \textit{end} node.

@@

\subsection{Node transformation}
Nodes do not require to be specified using a separate node specification rule, and will simply be generated in case they are used within an edge. Node specifications function specifically for the cases in which additional attributes need to be added to nodes. To support operations as well as general labels, the only two attributes which will be tracked from node specification are the \textit{label} and \textit{xlabel} arguments. These labels will be matched with regexes specifying the supported formats for operations and conditions. In case the label does not match either, it is still tracked as regular node labels are still relevant for reachability reporting. For the approach discussed within this paper the node labels represent the code line from which they were generated and are therefore needed to analyse code line reachability. All node specifications will follow the format given below where the square bracket can contain a number of optional arguments.

@@

\subsection{Initial node}
For reachability analysis to be relevant there needs to be an initial node. In automata, the initial node is often represented by a transition going from nowhere towards a single node. This is however not a built in feature within the dot language. One of the solutions to overcome this is to create a node and make it invisible, an edge can then be drawn from this invisible node to the actual initial node. Other approaches are potentially possible but were not explored. Any alternative implementations could easily be introduced instead of the considered format without interchanging any of the other components.

@@

\section{Reachability of non-parametric COCA}
\label{sec:reachability non parametric}
In this section we will give a description of the solution that was used to determine the reachability for non-parametric continuous one counter automata. This will be done by showing the approach that was taken, as well as an application to show its usage. This approach is a practical implementation of the theoretical approach described by Michael Blondin et al. \cite{blondin2021continuous}.

\subsection{Initialization}
During the initialization phase, we will instantiate all reachability intervals. The initial node will get an initial reachability interval $[i, i]$ where $i$ represents the initial counter value. This is not a fixed value, but will for the sake of this research always default to 0, simply because all integer values in most programming languages default to 0 at initialization.

@@

\subsection{Interval updates}
\label{interval updates}
From the initial state, we will start doing cycles in which we update all intervals once, wherever possible. This will be done using the successor function \textit{succ} below where $R_i: Q \times Q \rightarrow 2^\mathbb{Q}$ represent the mapping function for the i-th iteration. This function is going from each transition to its corresponding counter reachability interval. The \textit{succ} function will update the interval of one transition, \textit{(p, z, q)}, to its next iteration, based upon the current configuration.

@@

\subsection{Loop acceleration}
\label{sec:accel}
Unfortunately, the \textit{succ} function is not guaranteed to find the bounds in a finite number of evaluations. For example, the scenario given in Figure 23 would never converge at all. The interval of the loop between $q_1$ and $q_2$ will keep expanding indefinitely and the evaluation will therefore never stop.

@@

\subsection{Proof of termination}
The mathematical proof of termination has already been given by Michal Blondin et al. \cite{blondin2021continuous} but an intuitive proof for the algorithm will also be given here. Given that there are no loops it will trivially terminate within n evaluations, given that n is equal to the total number of nodes within the automaton. If there are loops they must either be positively expanding, negatively expanding, expanding in both directions or not expanding at all.

@@

\section{Reachability of parametric COCA}
\label{sec:reachability parametric}
Throughout this section we will give an overview of how the reachability for parametric continuous one counter automata can be analysed. This will be done using the approach that was theoretically described by Michael Blondin et al. \cite{blondin2021continuous}. This paper will both give a practical approach as well as an application, intended as prove that it works in practise.

@@

\subsection{Transition initialization}

@@

\subsection{Node condition initialization}

@@

\subsection{Interval encoding}

@@

\subsection{Define the end goal}
One by one we will verify for each node whether or not it is reachable, this can be done using the condition below, where e represents the edge taken to generate any of the intervals part of the goal node. In case e is equal to -2 it signifies that no transition was taken and that the interval is empty. If e is not equal to -2 it must imply that the according interval is not empty as will be explained later.

@@

\subsection{msum}
The msum function is used to determine the inclusiveness of the bounds in the interval addition. The msum function is defined as:

@@

\subsection{Emptiness of intervals}
We need to be able to verify whether or not an interval is empty in order to create base cases for future operations. An interval is considered to be empty if both bounds are not inclusive and the boundary values are equal to each other as can be seen in the formulae below.

@@

\subsection{Interval addition}
The addition of intervals will be split in three different cases as can be seen in the graph below. In case either operand is an empty interval the resulting interval will trivially equal the other interval. In case they are both not empty we compute the sum by taking the addition of the lower and upper bound of the X and Y intervals and the msum to compute whether or not the resulting bounds are inclusive. An overview of this can be found in Figure \ref{fig:addition}.

@@

		\draw [arrow] (q1) -> (q4);
		\draw [arrow] (q2) -> (q5);
		\draw [arrow] (q3) |- (q6);
	\end{tikzpicture}
	\caption{Overview of the interval addition operation}
	\label{fig:addition}
\end{figure}

\subsection{Interval overlap}
\begin{figure}[b!]
	\centering
	\begin{tikzpicture}[auto, >=latex, font=\footnotesize\scshape]
		\tikzstyle{small} = [draw=black, rectangle]
		\tikzstyle{center} = [draw=black, rectangle, text width=8em, align=center]
		\tikzstyle{arrow} = [thick,->,>=stealth]

@@

\subsection{Interval union}
The union operation is summarised in Figure \ref{fig:union} below. We can immediately see that this is very similar to the intersection in the sense that most of the conditional statements are the exact same.
